Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-03-27 Delaneyetal2024.tex
Updates from Overleaf
small changes with grammer/formating

\correspondingauthor{Ian Delaney}{ianarburua.delaney@unil.ch}

\begin{keypoints}
\item Short water discharge fluctuations in subglacial channels are accommodated by water velocity, not channel size, as in subaerial systems.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Significant subglacial sediment transport capacity can persist across a range of water discharges, impacting the  timing of peak events.
\end{keypoints}

@@

Over long periods, such as millenia, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially.

@@

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \cite{leopold1953}.
The change in water depth results in a commensurate increase in water velocity.
As a result, the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

@@

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Models of subglacial and subaerial hydraulics are used to quantify the respective water velocity and channel morphology.
The models are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The  lumped nature of the models isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in records of sediment transport, and their relationship with hydrology from glacierized catchments.

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming no water storage in the proglacial area.
The Alpine site (labeled \alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014 to October 10, 2014 \citeA{felix2021,felix2022}.

>>>>

\correspondingauthor{Ian Delaney}{ianarburua.delaney@unil.ch}
\ian{quick check on shear stress units: seems like Pa, not Pa/m}
\begin{keypoints}
\item Short water discharge fluctuations in subglacial channels are accommodated by water velocity, not channel size, as in subaerial systems.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Significant subglacial sediment transport capacity can persist across a range of water discharges, impacting the  timing of peak events.
\end{keypoints}

@@

Over long periods, such as millennia, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially.

@@

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \cite{leopold1953}.
The change in water depth results in a proportional increase in water velocity.
As a result, the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

@@

This manuscript evaluates the relationship between sediment transport capacity and water discharge in subglacial channels and compares it to subaerial ones.
Models of subglacial and subaerial hydraulics are used to quantify the respective water velocity and channel morphology.
The models are applied to hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and  a land-terminating glacier in Greenland (Leverett Glacier).
The  lumped nature of the models isolates the disparity between water discharge and sediment transport capacity in subglacial systems, independent of the upstream drainage network and sediment access.
Outputs demonstrate differences in the relationships between  water discharge, channel geometry, and water velocity in the subglacial and subaerial channels.
The manuscript then uses the demonstrated differences between the two channel types to discuss the implications for interpreting the processes driving variability in sediment transport records and their relationship with hydrology from glacierized catchments.

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming no water storage in the proglacial area.
The Alpine site (labeled \alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014, to October 10, 2014 \cite{felix2021,felix2022}.
